---
title: "Ensembles Lab 2: Boosting"
authors:
- Adapted by EVL, FRC and ASP
date: "`r Sys.Date()`"
format:
    html: 
      toc: true
      toc-depth: 3
      code-fold: false
      fig-width: 8
      fig-height: 6
knit:
  quarto:
    chunk_options:
      echo: true
      cache: false
      prompt: false
      tidy: true
      comment: NA
      message: false
      warning: false
    knit_options:
      width: 75
reference-location: margin
execute:
    echo: true
    message: false
    warning: false
    cache: true
bibliography: "StatisticalLearning.bib"
editor_options: 
  chunk_output_type: console
---

```{r packages, include=FALSE}
# If the package is not installed then it will be installed
if(!require(rsample)) install.packages("rsample", dep=TRUE)
if(!require("modeldata")) install.packages("modeldata")
# if(!require("tidymodels")) install.packages("tidymodels")
if(!require("xgboost")) install.packages("xgboost")
if(!require("gbm")) install.packages("gbm")
```

```{r message=FALSE}
# Helper packages
library(dplyr)       # for data wrangling
library(ggplot2)     # for awesome plotting
library(modeldata)  
library(foreach)     # for parallel processing with for loops

# Modeling packages
# library(tidymodels)
library(xgboost)
library(gbm)
```

# Introduction

This lab continues on the previous one showing how to apply boosting. The same dataset as before will be used


## Ames Housing dataset

Packge `AmesHousing` contains the data jointly with some instructions to create the required dataset.

We will use, however data from the `modeldata` package where some preprocessing of the data has already been performed (see: [https://www.tmwr.org/ames](https://www.tmwr.org/ames))


The dataset has 74 variables so a descriptive analysis is not provided.

```{r}
dim(ames)
boxplot(ames)
```

We proceed as in the previous lab and divide the reponse variable by 1000 facilitate reviewing the results .

```{r}
require(dplyr)
ames <- ames %>% mutate(Sale_Price = Sale_Price/1000)
boxplot(ames)
```

## Spliting the data into test/train

The data are split in separate test / training sets and do it in such a way that samplig is balanced for the response variable, `Sale_Price`.

```{r}
# Stratified sampling with the rsample package
set.seed(123)
split <- rsample::initial_split(ames, prop = 0.7, 
                       strata = "Sale_Price")
ames_train  <- training(split)
ames_test   <- testing(split)
```

# Fitting a boosted regression tree with `xgboost`

## XGBoost Parameters Overview

The `xgboost()` function in the XGBoost package trains Gradient Boosting models for regression and classification tasks. Key parameters and hyperparameters include:

- **Parameters**:
  - `params`: List of training parameters.
  - `data`: Training data.
  - `nrounds`: Number of boosting rounds.
  - `watchlist`: Validation set for early stopping.
  - `obj`: Custom objective function.
  - `feval`: Custom evaluation function.
  - `verbose`: Verbosity level.
  - `print_every_n`: Print frequency.
  - `early_stopping_rounds`: Rounds for early stopping.
  - `maximize`: Maximize evaluation metric.
  - `save_period`: Model save frequency.
  - `save_name`: Name for saved model.
  - `xgb_model`: Existing XGBoost model.
  - `callbacks`: List of callback functions.
  
### XGBoost Parameters Overview

Numerous parameters govern XGBoost's behavior. A detailed description of all parameters can be found in the `XGBoost` documentation. Key considerations include those controlling tree growth, model learning rate, and early stopping to prevent overfitting:

- **Parameters**:
  - `booster` [default = gbtree]: Type of weak learner, trees ("gbtree", "dart") or linear models ("gblinear").
  - `eta` [default=0.3, alias: learning_rate]: Reduces each tree's contribution by multiplying its original influence by this value.
  - `gamma` [default=0, alias: min_split_loss]: Minimum cost reduction required for a split to occur.
  - `max_depth` [default=6]: Maximum depth trees can reach.
  - `subsample` [default=1]: Proportion of observations used for each tree's training. If less than 1, applies Stochastic Gradient Boosting.
  - `colsample_bytree`: Number of predictors considered at each split.
  - `nrounds`: Number of boosting iterations, i.e., the number of models in the ensemble.
  - `early_stopping_rounds`: Number of consecutive iterations without improvement to trigger early stopping. If NULL, early stopping is disabled. Requires a separate validation set (watchlist) for early stopping.
  - `watchlist`: Validation set used for early stopping.
  - `seed`: Seed for result reproducibility. Note: use `set.seed()` instead.

## Test / Training in xGBoost

### XGBoost Data Formats

XGBoost models can work with various data formats, including R matrices. 

However, it's advisable to use `xgb.DMatrix`, a specialized and optimized data structure within this library.

```{r}
ames_train <- xgb.DMatrix(
                data  = ames_train %>% select(-Sale_Price)
                %>% data.matrix(),
                label = ames_train$Sale_Price,
               )

ames_test <- xgb.DMatrix(
                data  = ames_test %>% select(-Sale_Price)
                %>% data.matrix(),
                label = ames_test$Sale_Price
               )
```

## Fit the model

```{r fitBoost1}
set.seed(123)
ames.boost <- xgb.train(
            data    = ames_train,
            params  = list(max_depth = 2),
            nrounds = 1000,
            eta= 0.05
          )
ames.boost
```

## Prediction and model assessment

```{r}
ames.boost.trainpred <- predict(ames.boost,
                   newdata = ames_train
                 )

ames.boost.pred <- predict(ames.boost,
                   newdata = ames_test
                 )

train_rmseboost <- sqrt(mean((ames.boost.trainpred - getinfo(ames_train, "label"))^2))

test_rmseboost <- sqrt(mean((ames.boost.pred - getinfo(ames_test, "label"))^2))

paste("Error train (rmse) in XGBoost:", round(train_rmseboost,2))
paste("Error test (rmse) in XGBoost:", round(test_rmseboost,2))
```


# Parameter optimization

Extra lab "labs/Lab-C.2.4b-BoostingOptimization.qmd" shows how to optimize some important parameters for boosting.


<!-- ## Tree number -->

<!-- This is a critical parameter as far as adding new trees increases risk of overfitting. -->

<!-- Before optimization is run, data is shaped into an object of class `xgb.DMatrix`, which is required to run XGBoost through this package. -->

<!-- ```{r} -->
<!-- ames_train_num <- model.matrix(Sale_Price ~ . , data = ames_train)[,-1] -->
<!-- ames_test_num <- model.matrix(Sale_Price ~ . , data = ames_test)[,-1] -->

<!-- train_labels <- ames_train$Sale_Price -->
<!-- test_labels <- ames_test$Sale_Price -->

<!-- ames_train_matrix <- xgb.DMatrix( -->
<!--   data = ames_train_num, -->
<!--   label = train_labels -->
<!-- ) -->

<!-- ames_test_matrix <- xgb.DMatrix( -->
<!--   data = ames_test_num, -->
<!--   label = test_labels -->
<!-- ) -->

<!-- ``` -->


<!-- ```{r} -->
<!-- boostResult_cv <- xgb.cv( -->
<!--   data = ames_train_matrix, -->
<!--   params = list(eta = 0.3, max_depth = 6, subsample = 1, objective = "reg:squarederror"), -->
<!--   nrounds = 500, -->
<!--   nfold = 5, -->
<!--   metrics = "rmse",  -->
<!--   verbose = 0 -->
<!-- ) -->

<!-- boostResult_cv <- boostResult_cv$evaluation_log -->
<!-- print(boostResult_cv) -->
<!-- ``` -->

<!-- We aim at at the lowest number of trees that has associated a small cross-validation error. -->


<!-- ```{r} -->
<!-- ggplot(data = boostResult_cv) + -->
<!--   geom_line(aes(x = iter, y = train_rmse_mean, color = "train rmse")) +  -->
<!--   geom_line(aes(x = iter, y = test_rmse_mean, color = "cv rmse")) + -->
<!--   geom_point( -->
<!--     data = slice_min(boostResult_cv, order_by = test_rmse_mean, n = 1), -->
<!--     aes(x = iter, y = test_rmse_mean), -->
<!--     color = "firebrick" -->
<!--   ) + -->
<!--   labs( -->
<!--     title = "Evolution of cv-error vs number of trees", -->
<!--     x     = "number of trees", -->
<!--     y     = "cv-error (rmse)", -->
<!--     color = "" -->
<!--   ) + -->
<!--   theme_bw() + -->
<!--   theme(legend.position = "bottom") -->
<!-- ``` -->

<!-- ```{r} -->
<!-- paste("Optimal number of rounds (nrounds):", slice_min(boostResult_cv, order_by = test_rmse_mean, n = 1)$iter) -->
<!-- ``` -->

<!-- ## Learning rate -->

<!-- Alongside the number of trees, the learning rate (eta) is the most crucial hyperparameter in Gradient Boosting. It controls how quickly the model learns and thus influences the risk of overfitting.  -->

<!-- These two hyperparameters are interdependent: a lower learning rate requires more trees to achieve good results but reduces the risk of overfitting. -->

<!-- ```{r} -->
<!-- # Rango de valores para la tasa de aprendizaje (eta) -->
<!-- eta_range <- c(0.001, 0.01, 0.1, 0.3) -->
<!-- df_results_cv <- data.frame() -->

<!-- for (i in seq_along(eta_range)) { -->
<!--   set.seed(123) -->

<!--   # Validación cruzada con el eta actual -->
<!--   results_cv <- xgb.cv( -->
<!--     data = ames_train_matrix,  # ✅ Usamos el xgb.DMatrix correcto -->
<!--     params = list( -->
<!--       eta = eta_range[i],  -->
<!--       max_depth = 6,  -->
<!--       subsample = 1,  -->
<!--       objective = "reg:squarederror" -->
<!--     ), -->
<!--     nrounds = 1000, -->
<!--     nfold = 5, -->
<!--     metrics = "rmse",  -->
<!--     verbose = 0 -->
<!--   ) -->

<!--   # Extraer la evaluación de RMSE y registrar resultados -->
<!--   results_cv <- results_cv$evaluation_log -->
<!--   results_cv <- results_cv %>% -->
<!--     select(iter, test_rmse_mean) %>% -->
<!--     mutate(eta = as.character(eta_range[i]))  # Guardamos el eta usado -->

<!--   df_results_cv <- df_results_cv %>% bind_rows(results_cv) -->
<!-- } -->

<!-- ``` -->


<!-- ```{r} -->
<!-- ggplot(data = df_results_cv) + -->
<!--   geom_line(aes(x = iter, y = test_rmse_mean, color = eta)) +  -->
<!--   labs( -->
<!--     title = "Evolución del error en validación cruzada vs tasa de aprendizaje (eta)", -->
<!--     x = "Número de iteraciones", -->
<!--     y = "Error RMSE en validación cruzada", -->
<!--     color = "Eta" -->
<!--   ) + -->
<!--   theme_bw() + -->
<!--   theme(legend.position = "bottom") -->
<!-- ``` -->


<!-- ## Optimized predictor -->

<!-- In order to obtain improved predictors one can perform a a grid search for the best parameter combination can be performed. -->

<!-- ```{r prepareOptimization} -->

<!-- # Convertir variables categóricas a dummy variables usando model.matrix() -->
<!-- ames_train_num <- model.matrix(Sale_Price ~ . , data = ames_train)[,-1] -->
<!-- ames_test_num <- model.matrix(Sale_Price ~ . , data = ames_test)[,-1] -->

<!-- # Extraer etiquetas de Sale_Price -->
<!-- train_labels <- ames_train$Sale_Price -->
<!-- test_labels <- ames_test$Sale_Price -->

<!-- # Convertir a xgb.DMatrix -->
<!-- ames_train_matrix <- xgb.DMatrix( -->
<!--   data = ames_train_num, -->
<!--   label = train_labels -->
<!-- ) -->

<!-- ames_test_matrix <- xgb.DMatrix( -->
<!--   data = ames_test_num, -->
<!--   label = test_labels -->
<!-- ) -->

<!-- # Range of parameter values to test -->
<!-- eta_values <- c(0.01, 0.05, 0.1, 0.3)   -->
<!-- nrounds_values <- c(500, 1000, 2000)   -->

<!-- best_rmse <- Inf -->
<!-- best_params <- list() -->
<!-- ``` -->


<!-- ```{r runGridSearch} -->
<!-- cv_results_df <- data.frame() -->

<!-- set.seed(123) -->

<!-- for (eta in eta_values) { -->
<!--   for (nrounds in nrounds_values) { -->

<!--     cv_results <- xgb.cv( -->
<!--       data = ames_train_matrix,   -->
<!--       params = list( -->
<!--         eta = eta, -->
<!--         max_depth = 6, -->
<!--         subsample = 0.8, -->
<!--         colsample_bytree = 0.8, -->
<!--         objective = "reg:squarederror" -->
<!--       ), -->
<!--       nrounds = nrounds, -->
<!--       nfold = 5, -->
<!--       metrics = "rmse", -->
<!--       verbose = 0, -->
<!--       early_stopping_rounds = 10 -->
<!--     ) -->

<!--     if (is.null(cv_results)) next -->

<!--     results_row <- data.frame( -->
<!--       eta = eta, -->
<!--       nrounds = nrounds, -->
<!--       min_rmse = min(cv_results$evaluation_log$test_rmse_mean), -->
<!--       best_nrounds = cv_results$evaluation_log$iter[which.min(cv_results$evaluation_log$test_rmse_mean)] -->
<!--     ) -->

<!--     cv_results_df <- bind_rows(cv_results_df, results_row) -->

<!--     if (results_row$min_rmse < best_rmse) { -->
<!--       best_rmse <- results_row$min_rmse -->
<!--       best_params <- list( -->
<!--         eta = results_row$eta, -->
<!--         nrounds = results_row$best_nrounds -->
<!--       ) -->
<!--     } -->
<!--   } -->
<!-- } -->
<!-- ``` -->


<!-- ```{r showGridSearch} -->
<!-- cat("\n, Best hyperparameters values found:\n") -->
<!-- cat("Eta:", best_params$eta, "\n") -->
<!-- cat("Nrounds:", best_params$nrounds, "\n") -->
<!-- cat("RMSE mínimo:", round(best_rmse, 4), "\n") -->

<!-- ``` -->

# References
