---
title: "Prediction with (shallow) Neural Networks"
authors:
- Adapted by EVL, FRC and ASP
date: "`r Sys.Date()`"
format:
    html: 
      toc: true
      toc-depth: 3
      code-fold: show
      fig-width: 8
      fig-height: 6
knit:
  quarto:
    chunk_options:
      echo: true
      cache: false
      prompt: false
      tidy: true
      comment: NA
      message: false
      warning: false
    knit_options:
      width: 75
reference-location: margin
execute:
    echo: true
    message: false
    warning: false
bibliography: "../StatisticalLearning.bib"
editor_options: 
  chunk_output_type: console
---

# Building a NN with the `neuralnet` package

## A predictive ANN

We use the `neuralnet` package to build a simple neural network to predict if a type of stock pays dividends or not.

```{r echo=TRUE}
if (!require(neuralnet)) 
  install.packages("neuralnet", dep=TRUE)
```

## Data for the example

And use the `dividendinfo.csv` dataset from <https://github.com/MGCodesandStats/datasets>

```{r echo=TRUE}
mydata <- read.csv("https://raw.githubusercontent.com/MGCodesandStats/datasets/master/dividendinfo.csv")
str(mydata)
```

## Data pre-processing

```{r echo=TRUE}
normalize <- function(x) {
  return ((x - min(x)) / (max(x) - min(x)))
}
normData <- as.data.frame(lapply(mydata, normalize))
```

## Test and training sets

Finally we break our data in a test and a training set:

```{r echo=TRUE}
perc2Train <- 2/3
ssize <- nrow(normData)
set.seed(12345)
data_rows <- floor(perc2Train *ssize)
train_indices <- sample(c(1:ssize), data_rows)
trainset <- normData[train_indices,]
testset <- normData[-train_indices,]
```

## Training a neural network

We train a simple NN with two hidden layers, with 4 and 2 neurons respectively.

First we deine the parameters and hyperparameters

```{r}
hidden_layers <- c(4,2)  # Two hidden layers with 4 and 2 neurons
activation_function <- "logistic"  # Sigmoid
optimizer <- "rprop+"  # Optimizer
learning_rate <- 0.01  # Learning rate
threshold_value <- 0.01  # Convergence  thresholod 
max_iterations <- 1e6  # Max iterations
```


```{r echo=TRUE}
# Neural Network
library(neuralnet)
nn <- neuralnet(#dividend ~ .,
                dividend ~ fcfps + earnings_growth + de + mcap + current_ratio, 
                data=trainset, 
                hidden=hidden_layers, 
                act.fct=activation_function, 
                algorithm=optimizer, 
                learningrate=learning_rate,
                linear.output=FALSE, 
                threshold=threshold_value,
                stepmax=max_iterations)
```

## Network plot

The output of the procedure is a neural network with estimated weights.

```{r}
plot(nn, rep = "best")
```


In order to facilitate visualization we can round decimals to 2 for all weights.

```{r echo=TRUE}
round_weights <- function(nn_model, decimals=2) {
  nn_model$weights <- lapply(nn_model$weights, function(layer) {
    lapply(layer, function(weight_matrix) {
      if (is.numeric(weight_matrix)) {
        round(weight_matrix, decimals)
      } else {
        weight_matrix
      }
    })
  })
  return(nn_model)
}

# Apply function to trained net
nn_rounded <- round_weights(nn, 2)

# View network with less decimals
plot(nn_rounded, rep = "best")

```

## Predictions

```{r echo=TRUE}
temp_test <- subset(testset, select =
                      c("fcfps","earnings_growth", 
                        "de", "mcap", "current_ratio"))
nn.results <- compute(nn, temp_test)
results <- data.frame(actual = 
                  testset$dividend, 
                  prediction = nn.results$net.result)
head(results)
```

## Model evaluation

```{r echo=TRUE}
roundedresults<-sapply(results,round,digits=0)
roundedresultsdf=data.frame(roundedresults)
attach(roundedresultsdf)
table(actual,prediction)
```

The `caret` package can be used to provide a better confusion matrix.

```{r}
# Round predictions to obtain binary values (0 or 1)
roundedresults <- sapply(results, round, digits=0)
roundedresultsdf <- data.frame(roundedresults)

# Convert variables to factors for proper processing in `caret`
roundedresultsdf$actual <- as.factor(roundedresultsdf$actual)
roundedresultsdf$prediction <- as.factor(roundedresultsdf$prediction)

# Load caret package (install if not available)
if (!require(caret)) install.packages("caret", dependencies=TRUE)
library(caret)

# Generate confusion matrix with advanced metrics
conf_matrix <- confusionMatrix(roundedresultsdf$prediction, roundedresultsdf$actual)

# Print the confusion matrix and performance metrics
print(conf_matrix)

```

# Full workflow using `Caret`

The `Caret` package offers an alternative to `neuralnet` to perform a more efficient process.

```{r}
# Load required libraries (install if necessary)
if (!require(caret)) install.packages("caret", dependencies=TRUE)
if (!require(nnet)) install.packages("nnet", dependencies=TRUE)  # Needed for neural networks
library(caret)
library(nnet)
```

Data can be normalized using using caret's `preProcess` function

```{r}
preproc <- preProcess(mydata, method = c("range"))  # Min-Max normalization
normData <- predict(preproc, mydata)
```

Test/Train splitting is automated

```{r}
# Split data into training (67%) and testing (33%) sets
set.seed(12345)
trainIndex <- createDataPartition(normData$dividend, p = 0.67, list = FALSE)
trainset <- normData[trainIndex,]
testset <- normData[-trainIndex,]
```

Model variability can be managed by crossvalidation

```{r}
ctrl <- trainControl(method = "cv", number = 10)  # 10-fold cross-validation
```


```{r}
# Train a neural network model using caret's `train` function
set.seed(12345)
nn_model <- train(dividend ~ ., data = trainset,
                  method = "nnet",  # Use neural network model
                  trControl = ctrl,  # Cross-validation
                  tuneGrid = expand.grid(size = c(4, 2), decay = 0.01),  # Hidden layers, weight decay
                  maxit = 1000,  # Max iterations
                  trace = FALSE)  # Suppress output during training

# Print model summary
print(nn_model)
```
Once the model is trained we can proceed as usually

```{r}
# Make predictions on test set
y_pred <- predict(nn_model, testset)

# Convert predictions to binary (0 or 1)
y_pred_bin <- ifelse(y_pred > 0.5, 1, 0)

# Convert actual values and predictions to factors for confusion matrix
testset$dividend <- as.factor(testset$dividend)
y_pred_bin <- as.factor(y_pred_bin)

# Generate confusion matrix and performance metrics
conf_matrix <- confusionMatrix(y_pred_bin, testset$dividend)

# Print confusion matrix and evaluation metrics
print(conf_matrix)

```

Notice that the model shows slightly better than the first one, probably due to the fact that it has been minimally optimized.